A Study of Residual Adapters for Multi-Domain Neural Machine Translation

### 2020-10-06 2020/wmt_20/mdadapt.tex
reviewed till page 7

%\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\}

@@

Under this view, building adapted systems is a two-step process: (a) one first trains NMT with the largest possible parallel corpora, possibility aggregating texts from multiple, heterogeneous sources; (b) assuming that in-domain parallel documents are available for the domain of interest, one then adapts the pre-trained model by resuming training with the sole in-domain corpus. It is a conjecture that the pretrained model constitutes a better initialization than a random one, especially when adaptation data is scarce. Indeed, studies of transfer learning for NMT such as \citet{artetxe20cross,aji20neural} have confirmed this claim in extensive experiments. Full fine-tuning, that adapts all the parameters of a baseline model usually significantly improves the quality of the NMT for the chosen domain. However, it also yields large losses in translation quality for other domains, a phenomenon referred to as ``catastrophic forgetting'' in the neural network literature \cite{McCloskey89catastrophic}. Therefore, a fully fine-tuned model is \emph{only useful to one target domain}. As the number of domains to handle grows, training, and maintaining a separate model for each task can quickly become tedious and resource-expensive .\fyDone{Fix this sentence.}

\citet{Vilar18learning,Bapna19simple,wuebker18compact,michel18extreme} have recently proposed a simple and lightweight domain adaptation method that also preserves the value of pre-trained models, based on small adapter components that are plugged in each hidden layer. These adapters are trained only with the in-domain data, keeping the pre-trained model frozen. Because these additional adapters are very small compared to the size of the baseline model, their use significantly reduces the cost of training and maintaining fine-tuned models, while still delivering a performance that is close to that of full fine-tuning.

@@

\section{Residual adapters \label{sec:res}}
In this section, we describe the basic version of the residual adapter architectures \citep{houlsby19parameter, Bapna19simple}, as well as two original variants of this model.

@@

\subsubsection{The computation of adapter layers}
Our reference architecture is the Transformer model of \citet{Vaswani17attention}, which we assume contains a stack of layers both on the encoder and the decoder sides. Each layer contains two subparts, an attention layer, and a dense layer. Details vary from one implementation to another, we simply contend here that each layer $i \in \{1 \dots L\}$ (in the encoder or the decoder) computes a transform of a fixed-length sequence of $d$-dimensional input vectors $h^{i}$ into a sequence of output vectors $h^{i+1}$.

The residual adapter architecture extends this model by adding one new computation module for each layer. The $i^{\text{th}}$ residual adapter thus modifies $h^i$ by computing the following transformations:\fyDone{Use align env} \fyTodo{Add layer normalization, replace notation b by a}
\begin{align*}
  h^{i}_0 &= \mathbf{LN}(h^{i}) \\
  h^{i}_1 &= \mathbf{W}_{db}^{i}h_0^{i} + a^i_{1} \\
  h^{i}_2 &= \mathbf{ReLU}(h_1^{i})
  \\
%\end{align*}
%\begin{align*}
  h^{i}_3 &= \mathbf{W}_{bd}^{i}h_2^{i} + a^i_{2} \\
  \bar{h}^{i} &= h^{i}_3 + h^i.
\end{align*}
Overall, the  $i^{\text{th}}$ adapter is thus parameterized by matrices $\displaystyle{\mathbf{W}_{db}^{i}\in\mathbb{R}^{d\times b}}$,$\displaystyle{\mathbf{W}_{bd}^{i}\in\mathbb{R}^{b\times d}}$, bias vectors $\displaystyle{b^i_{1} \in \mathbb{R}^{b}}$, $\displaystyle{b^i_{2} \in \mathbb{R}^{d}}$, with $b$ the dimension of the adapter \fyTodo{($d \gg b$) not true}\fyDone{Check this}. For the sake of brevity, we will simply denote $h^{i}_3 = \operatorname{ADAP}^{(i)}(h^i)$, and $\theta_{\operatorname{ADAP}^{(i)}}$ the corresponding set of parameters.\fyDone{or is it $h_i$ ?}\fyDone{attention aux matrices $W_i$}

@@

In the experiments of Section~\ref{sec:exp}, we use $2\times{}L=12$ residual adapters, one for each of the $L=6$ attention layers of the encoder and similarly for the decoder.\footnote{In the decoder, the stack of self-attention layers and cross encoder-decoder attention only counts as one attention layer and only serves one residual adapter.}

@@

The first question concerns the number of adapter layers. While in principle, all Transformer layers can be subject to adaptation, it is nonetheless worthwhile to consider simpler adaptation schemes, which would only alter a limited number of layers. Such strategy might be especially relevant when the training data contains very small domains, as in the experiments of Section~\ref{sec:exp}, and for which a complete adaptation may not be necessary or/and or prone to overfitting. Likewise, it might be meaningful to explore ways to share subsets of adapters across domains. This in turn raises the issue of which layer(s) to adapt, a question that can be approached in the light of recent analyses of Transformers models, which conjecture that the higher layers encode global patterns with a more ``semantic'' interpretation, while the lower layers encode local patterns akin to morpho-syntactic information \cite{raganato18analysis}.

@@

One obvious benefit of this variant is that it allows us to reuse the hidden vectors $h^i$ of all hidden layers when computing an adapted output for several domains during the inference. In this situation, the forward step needs only to compute the hidden vectors $h^i$ once for the inner encoder layers, before an adapted sequence of vectors is computed at the topmost layer. Therefore, we can finetune the model to multiple domains at once without recomputing $h^i$. This variant also opens the way to more parameter sharing across adapters, a perspective that we will not explore further in this work. Instead, we use it to develop a second variation of the adapter model, that is presented in the next section.

@@

Therefore, we would like to design a version of residual adapters that is more robust to such domain errors. This variant, called the \emph{gated residual adapter model}, relies on the training of a supplementary trained component that will help decide whether to activate, on a word per word basis, a given residual layer and to regulate the strength of this activation. To this end, we extend the highway version of residual adapters as follows.
\fyDone{Consistency of notations wrt section 2.1}

@@

\subsection{Baseline architectures \label{ssec:baseline}}
\fyDone{Write this - settings and parameters for Mixed and Full-FT}
% Our baselines are standard for multi-domain systems.
% \footnote{We however omit domain-specific systems trained only with the corresponding subset of the data, which are always inferior to the mix-domain strategy \cite{Britz17mixing}.}
Using Transformers \cite{Vaswani17attention} implemented in OpenNMT-tf\footnote{\url{https://github.com/OpenNMT/OpenNMT-tf}} \cite{Klein17opennmt}, we train the following baselines:
\begin{itemize}
\item a generic model trained on a concatenation of all corpora, denoted \system{Mixed};\fyDone{Or mixed nat ?}
\item a fine-tuned model \cite{Luong15stanford,Freitag16fast}, based on the \system{Mixed} system, further trained on each domain with early stopping when the dev BLEU stops increasing during 3 consecutive epochs.
%  We again contrast two versions: full fine-tuning (\system{FT-Full}), which update all the parameters of the initial generic model \system{Mixed}; and the variants of \cite{Bapna19simple} (\system{FT-Block}).
\end{itemize}

For all En$\rightarrow$Fr models, we set the embeddings size and the hidden layers size to~512. Transformers use multi-head attention with 8 heads in each of the 6 layers; the inner feedforward layer contains 2,048 cells. Residual adapters additionally use an adaptation block in each layer, composed of a 2-layer perceptron, with an inner ReLU activation function operating on normalized entries of dimension $b=1024$.\revision{ \citet{Bapna19simple} showed that the performance of adapted model increases with respect to the size of the inner dimension. In the experiments of \citet{Bapna19simple}, we note that $b=1024$, which is 2-folds larger than the dimension of Transformer layer, achieved performance closed to fine-tuned model. Therefore, we would like to apply this setting to our experiments.}
% The gated variant is made of a dense layer, followed by a layer normalization and a sigmoid activation.
% The domain control systems are exactly as their baseline counterparts (RNN and Transformer), with an additional 2 cells encoding the domain on the input layer.
Training uses a batch size of~12,288 tokens; optimization uses Adam with parameters $\beta_1=0.9$, $\beta_2= 0.98$ and Noam decay ($warmup\_steps=4,000$), and a dropout rate of $0.1$ for all layers.\revision{ For \system{mixed} model, we use initial learning 1.0 and the concatenation of the validation sets of 6 domains as the validation set of \system{mixed}. In the fine-tuning experiments, we continue training using \system{mixed} as starting point, we use the same learning rate schedule, continuing the incrementation of the number of steps. In the multi-task training, we use the same learning rate schedule as \system{mixed}, for each iteration, we sample a domain from the pool of 6 domains with the probability proportional to the size of training data of each domain, then we sample a batch of~12,288 tokens from training data of the picked domain, the shared parameters and the parameters of the corresponding adapter are updated by gradients.}\fyDone{Describe the block adaptation layer - voir slides}

Models for En$\rightarrow$De are larger and rely on embeddings size as well as hidden layers size of~1024; each Transformers layer contains 16~attention heads; the inner feedforward layer contains 4,096 cells. Adapter modules have the same architecture as for the other language pair, except for their size, which is doubled ($b=2,048$). 
%The number of parameters in each model is reported in Table~\ref{tab:params}.

@@

These results first show that full fine-tuning outperforms all other methods for the in-domain test sets. However, \system{Res-Adap} is able to reduce the gap with this approach for several domains, showing the effectiveness of residual adapters. The ``multi-task'' variant is slightly less effective in our experiments than the basic version, where optimization is performed in two steps. As it turns out, using residual adapters proves here more effective on average than the other reference multi-domain systems; it is also much better than the generic system for translating data from known domains, outperforming the \system{Mixed} system by more than 4 BLEU points in average. Gains are especially large for small domains such as \domain{law} and \domain{rel}.

@@

\subsection{Varying the positions and number of residual adapters}
Tables~\ref{tab:performance-en-fr-pos-reg} and \ref{tab:performance-en-de-pos-reg} report the BLEU scores for 6 domains in each language pair: \domain{med},\domain{law},\domain{bank},\domain{talk},\domain{it} and \domain{rel} for En$\rightarrow$Fr; \domain{gov}, \domain{eco}, \domain{tour}, \domain{bank}, \domain{med} and \domain{news} for En$\rightarrow$De. We first see that for the latter direction, the basic version \system{Res-Adap} also outperforms the \system{mixed} baseline on average, with large gains for the small domains \domain{tour}, \domain{bank} and comparable results for the other domains.
% in almost all domains while stays equivalent to the baseline in \domain{med} in En$\rightarrow$Fr and \domain{news}, \domain{gov}, \domain{eco}, \domain{med} in En$\rightarrow$De.
   %    However, \system{FT-full} still remains the best model for all test domains.
\fyFuture{Full FT for en->DE ? }

>>>>

\aclfinalcopy % Uncomment this line for the final submission
\def\aclpaperid{***} %  Enter the acl Paper ID here

@@

\author{MinhQuang Pham$^{\dag\ddag}$, \ \ Josep Crego$^\dag$,\ \ Fran\c cois Yvon$^\ddag$, \ \ Jean Senellart$^\dag$ \\
  $^\dag$SYSTRAN / 5 rue Feydeau, 75002 Paris, France\\
  {\tt firstname.lastname@systrangroup.com}\\
  $^\ddag$Universit\'e Paris-Saclay, CNRS, LIMSI,  91405 Orsay, France\\
  {\tt firstname.lastname@limsi.fr}}

@@

Under this view, building adapted systems is a two-step process: (a) one first trains NMT with the largest possible parallel corpora, aggregating texts from multiple, heterogeneous sources; (b) assuming that in-domain parallel documents are available for the domain of interest, one then adapts the pre-trained model by resuming training with the sole in-domain corpus. It is a conjecture that the pretrained model constitutes a better initialization than a random one, especially when adaptation data is scarce. Indeed, studies of transfer learning for NMT such as \citep{artetxe20cross,aji20neural} have confirmed this claim in extensive experiments. Full fine-tuning, that adapts all the parameters of a baseline model usually significantly improves the quality of the NMT for the chosen domain. However, it also yields large losses in translation quality for other domains, a phenomenon referred to as ``catastrophic forgetting'' in the neural network literature \cite{McCloskey89catastrophic}. Therefore, a fully fine-tuned model is \emph{only useful to one target domain}. As the number of domains to handle grows, training, and maintaining a separate model for each task can quickly become tedious and resource-expensive.\fyDone{Fix this sentence.}

Several recent studies (e.g.\ \citep{Vilar18learning,Wuebker18compact,Michel18extreme,Bapna19simple}) have proposed more lightweight schemes to perform domain adaptation, while also preserving  the value of pre-trained models. Our main inspiration is the latter work, whose proposal relies on small \emph{adapter components} that are plugged in each hidden layer. These adapters are trained only with the in-domain data, keeping the pre-trained model frozen. Because these additional adapters are very small compared to the size of the baseline model, their use significantly reduces the cost of training and maintaining fine-tuned models, while delivering a performance that remains close to that of full fine-tuning.

@@

\section{Residual adapters \label{sec:res}}
In this section, we describe the basic version of the residual adapter architectures \citep{houlsby19parameter, Bapna19simple}, as well as two novel variants of this model.

@@

\subsubsection{The computation of adapter layers}
Our reference architecture is the Transformer model of \citet{Vaswani17attention}, which we assume contains a stack of layers both on the encoder and the decoder sides. Each layer contains two subparts, an attention layer, and a dense layer. Details vary from one implementation to another, we simply contend here that each layer $i \in \{1 \dots L\}$ (in the encoder or the decoder) computes a transform of a fixed-length sequence of $d$-dimensional input vectors $h^{i}$ into a sequence of output vectors $h^{i+1}$ as follows ($\mathbf{LN}$ denotes the (sub)layer normalization, $\mathbf{ReLU}$ is the ``rectified linear unit'' operator):\fyDone{Use align env}\fyDone{Add layer normalization, replace notation b by a}
\begin{align*}
  h^{i}_0 &= \mathbf{LN}(h^{i}) \\
  h^{i}_1 &= \mathbf{W}_{db}^{i}h_0^{i} + a^i_{1} \\
  h^{i}_2 &= \mathbf{ReLU}(h_1^{i})
  \\
%\end{align*}
%\begin{align*}
  h^{i}_3 &= \mathbf{W}_{bd}^{i}h_2^{i} + a^i_{2} \\
  \bar{h}^{i} &= h^{i}_3 + h^i.
\end{align*}
Overall, the  $i^{\text{th}}$ adapter is thus parameterized by matrices $\displaystyle{\mathbf{W}_{db}^{i}\in\mathbb{R}^{d\times b}}$,$\displaystyle{\mathbf{W}_{bd}^{i}\in\mathbb{R}^{b\times d}}$, bias vectors $\displaystyle{b^i_{1} \in \mathbb{R}^{b}}$, $\displaystyle{b^i_{2} \in \mathbb{R}^{d}}$, with $b$ the dimension of the adapter \fyTodo{($d \gg b$) not true}\fyDone{Check this}. For the sake of brevity, we will simply denote $h^{i}_3 = \operatorname{ADAP}^{(i)}(h^i)$, and $\theta_{\operatorname{ADAP}^{(i)}}$ the corresponding set of parameters.\fyDone{or is it $h_i$ ?}\fyDone{attention aux matrices $W_i$}

@@

In the experiments of Section~\ref{sec:exp}, we use $2\times{}L=12$ residual adapters, one for each of the $L=6$ attention layers of the encoder and similarly for the decoder.\footnote{In the decoder, the stack of self-attention and cross encoder-decoder attention only counts as one attention layer and only corresponds to one residual adapter.}

@@

The first question is the number of adapter layers. While in principle, all Transformer layers can be subject to adaptation, it is nonetheless worthwhile to consider simpler adaptation schemes, which would only alter a limited number of layers. Such strategy might be especially relevant when the training data contains very small domains, as in the experiments of Section~\ref{sec:exp}, and for which a complete adaptation may not be necessary or/and or prone to overfitting. Likewise, it might be meaningful to explore ways to share subsets of adapters across domains. This, in turn, raises the issue of which layer(s) to adapt, a question that can be approached in the light of recent analyses of Transformers models, which conjecture that the higher layers encode global patterns with a more ``semantic'' interpretation, while the lower layers encode local patterns akin to morpho-syntactic information \cite{raganato18analysis}.

@@

One obvious benefit of this variant is that it allows us to reuse the hidden vectors $h^i$ of all hidden layers when computing an adapted output for several domains during the inference. In this situation, the forward step needs only to compute the hidden vectors $h^i$ once for the inner encoder layers, before an adapted sequence of vectors is computed at the topmost layer. Therefore, we can fine-tune the model to multiple domains at once without recomputing $h^i$. This variant also opens the way to more parameter sharing across adapters, a perspective that we will not explore further in this work. Instead, we use it to develop a second variation of the adapter model, that is presented in the next section.

@@

Therefore, we would like to design a version of residual adapters that is more robust to such domain errors. This variant, called the \emph{gated residual adapter model}, relies on the training of a supplementary component that will help decide whether to activate, on a word per word basis, a given residual layer and to regulate the strength of this activation. To this end, we extend the highway version of residual adapters as follows.
\fyDone{Consistency of notations wrt section 2.1}

@@

\subsection{Baseline architectures \label{ssec:baseline}}
\fyDone{Write this - settings and parameters for Mixed and Full-FT}
% Our baselines are standard for multi-domain systems.
% \footnote{We however omit domain-specific systems trained only with the corresponding subset of the data, which are always inferior to the mix-domain strategy \cite{Britz17mixing}.}
Using Transformers \cite{Vaswani17attention} implemented in OpenNMT-tf\footnote{\url{https://github.com/OpenNMT/OpenNMT-tf}} \cite{Klein17opennmt}, we train the following baselines:
\begin{itemize}
\item a generic model trained on a concatenation of all corpora, denoted \system{Mixed};\fyDone{Or mixed nat ?}
\item a fine-tuned model \cite{Luong15stanford,Freitag16fast}, based on the \system{Mixed} system, further trained on each domain with early stopping when the development BLEU score stops increasing during 3 consecutive epochs.
%  We again contrast two versions: full fine-tuning (\system{FT-Full}), which update all the parameters of the initial generic model \system{Mixed}; and the variants of \cite{Bapna19simple} (\system{FT-Block}).
\end{itemize}

For all En$\rightarrow$Fr models, we set the embeddings size and the hidden layers size to~512. Transformers use multi-head attention with 8 heads in each of the 6 layers; the inner feedforward layer contains 2,048 cells. Residual adapters additionally use an adaptation block in each layer, composed of a 2-layer perceptron, with an inner ReLU activation function operating on normalized entries of dimension $b=1024$. \revision{\citet{Bapna19simple} showed that the performance of adapted model increases with respect to the size of the inner dimension and obtained performance close to the full fine-tuned model with $b=1024$, which is twice as large as the dimension of a Transformer layer. We used the same setting in our experiments.}\fyDone{Check: is it $b$?}

% The gated variant is made of a dense layer, followed by a layer normalization and a sigmoid activation.
% The domain control systems are exactly as their baseline counterparts (RNN and Transformer), with an additional 2 cells encoding the domain on the input layer.
Training uses a batch size of~12,288 tokens; optimization uses Adam with parameters $\beta_1=0.9$, $\beta_2= 0.98$ and Noam decay ($warmup\_steps=4,000$), and a dropout rate of $0.1$ for all layers. \revision{For the \system{mixed} model, we use an initial learning rate of 1.0 and take the concatenation of the validation sets of 6 domains for development. In the fine-tuning experiments, we continue training using \system{mixed} as starting point, using the same learning rate schedule, and continuing the incrementation of the number of steps. In the multi-task training, we use the same learning rate schedule as for \system{mixed}: for each iteration, we sample a domain a probability proportional to its size; we then sample a batch of~12,288 tokens that is used to update the shared parameters and the parameters of the corresponding adapter.}\fyDone{Describe the block adaptation layer - voir slides}

Models for En$\rightarrow$De are larger and rely on embeddings as well as hidden layers of size~1024; each Transformers layer contains 16~attention heads; the inner feedforward layer contains 4,096 cells. Adapter modules have the same architecture as for the other language pair, except for their size, which is doubled ($b=2,048$).%\fyTodo{Check: is it $b$?}
%The number of parameters in each model is reported in Table~\ref{tab:params}.

@@

These results first show that full fine-tuning outperforms all other methods for the in-domain test sets. However, \system{Res-Adap} is able to reduce the gap with this approach for several domains, showing the effectiveness of residual adapters. The ``multi-task'' variant is slightly less effective in our experiments than the basic version, where optimization is performed in two steps. As it turns out, using residual adapters proves here better on average than the other reference multi-domain systems; it is also much better than the generic system for translating data from known domains, outperforming the \system{Mixed} system by more than 4 BLEU points in average. Gains are especially large for small domains such as \domain{law} and \domain{rel}.

@@

\subsection{Varying the positions and number of residual adapters}
Tables~\ref{tab:performance-en-fr-pos-reg}-\ref{tab:performance-en-de-pos-reg} report BLEU scores for 6 domains in each language pair: \domain{med},\domain{law},\domain{bank},\domain{talk},\domain{it} and \domain{rel} for En$\rightarrow$Fr; \domain{gov}, \domain{eco}, \domain{tour}, \domain{bank}, \domain{med} and \domain{news} for En$\rightarrow$De. We first see that for the latter direction, the basic version \system{Res-Adap} also outperforms the \system{mixed} baseline on average, with large gains for the small domains \domain{tour}, \domain{bank} and comparable results for the other domains.
% in almost all domains while stays equivalent to the baseline in \domain{med} in En$\rightarrow$Fr and \domain{news}, \domain{gov}, \domain{eco}, \domain{med} in En$\rightarrow$De.
   %    However, \system{FT-full} still remains the best model for all test domains.
\fyFuture{Full FT for en->DE ?}
